### 2021-07-06 rad-icfp-2021/intro.tex
.

In this paper, we define a purely functional language with efficient and provably correct reverse-mode AD.
To do so, we go start from a simple and general representation, where we can relatively easily prove correctness of reverse-mode.
Then, we define a representation that leads to more efficient implementations.

This raises several issues that we deal with throughout the paper. First, there is a tradeoff to reach between a general expressive language and a domain specific one. The latter usually has more static information and a specific representation that lends itself to better optimizations.
Then, many optimizations performed on AD implementations consist in hand-crafted derivatives for useful operations like matrix-matrix multiplication, dot-product, etc.
They don't seem to arise from theoretical justifications, are error prone, and can hardly fit with more general optimizations.
This makes these systems harder to prove correct. The problem is thus to ensure provable correctness and pureness, while not compromising on efficiency. In addition, we would like something easily provably efficient. 
A real-world implementation based on our work would of course optimize further, potentially using hand-crafted operations as well, but it would be based on solid grounds.

TODO: same as above, maybe rewrite for more theory audience.

\subsection{Summary of contributions}

We propose a source-code transformation on a simple purely functional language for purely functional reverse-mode differentiation.
Our transformation is comprised of a compilation scheme through a novel intermediate representation (IR): unary form (UNF).
To emphasize the key ideas and for expository purposes, we present our work with a simple yet expressive language in Section~\ref{sec:simplediff}, 
and show how to extend our work to a richer language in Section~\ref{sec:generalization}.

We summarize the key ideas of our work. Reverse-mode can be made efficient
\begin{itemize}
   \item easily if there are only unary operators
   \item in a purely functional way
   \item on expressive constructs such as second-order array operators (e.g. !map2, reduce!) 
   \item via standard functional optimization techniques
\end{itemize}

Our contribution makes these claims precise. In more details, we introduce
\begin{itemize}
    \item an expressive source language with array operations, yet restrictive enough for efficient differentiation (\S\ref{sub:sourcelang})
    \item a method for extending our work and modular approaches to add and prove correct new primitives as needed in practice (\S\ref{sec:generalization})
    \item a new intermediate representation (IR), UNF, essentially turning every operation into a unary one which inputs and outputs a tuple (\S\ref{sec:unf})
    \item a transformation between the IR and the main language (\S\ref{sub:transformations to and from UNF})
    \item a simple reverse-mode transformation on the IR (\S\ref{sub:Simple reverse mode transformation})
    \item the direct reverse-mode transformation obtained from the source to the target language (\S\ref{sub:Macro for pure reverse mode transformation})
    \item a denotational semantics of our languages and transformations using multicategories and concategories (\S\ref{sec:correctness})
    \item correctness of the transformations (\S\ref{sec:correctness})
    \item complexity guarantees (\S\ref{sec:complexity})
    \item how efficiency is ensured via standard optimization techniques (\S\ref{sub:Optimizations})
    \item several additions to the source language (\S\ref{sec:generalization}) TODO:should emphasize that it should be easier for us to add primitives compared to several recent work. give recipe.
\end{itemize}

% \begin{table}
% \begin{tabular}{|l|c|c|c|c|c|c|c|c|c|}
% \hline
%  & \rot{Reverse Mode}  & \rot{Complexity Proofs} & \rot{Pure} & \rot{Correctness Proofs} & \rot{Tensor Support}  & \rot{Recursion} & \rot{Conditional} \\
% \hline
% This Paper &
% \supfull & \supfull & \supfull  & \supfull  & \supfull  & \suphalf & \supfull \\ \hline
% Lantern~\cite{lantern_icfp} & 
% \supfull & \suphalf & \supnone  & \supnone & \supfull & ? & \supfull \\ \hline
% \dfsmooth{}~\cite{shaikhha2019efficient} & 
% \supnone & \supnone & \supfull  & \supnone & \supfull  & \suphalf  & \supnone \\ \hline
% \cite{huot2020correctness} &
% \supfull & \supnone & \supfull   & \supfull & \supnone & \suphalf & \supfull \\ \hline
% \cite{brunel2019backpropagation} &
% \supfull & \suphalf ? & \supfull & \supfull & \supnone & \supnone & ? \\ \hline
% \cite{abadi2019simple} &
% \supfull & ? & \supfull & \supfull & \supnone & \supfull & \supfull \\ \hline
% \cite{barthe2020versatility} &
% \supnone & \supnone & \supfull & \supfull & \supnone & \supnone & \supfull \\ \hline
% \cite{pearlmutter2008reverse} &
% \supfull & \suphalf & \supnone & \supnone & ? & ? & ? \\ \hline
% \cite{Elliott:2018:SEA:3243631.3236765} &
% \supfull & \suphalf & \supfull & \supnone & ? & \supnone & ? \\ \hline
% \end{tabular}
% \caption{Comparison of different functional differentiable programming frameworks.}
% \end{table}

\begin{table}
 \label{fig:comparison-table}
 \begin{tabular}{|l|c|c|c|c|c|c|c|c|}
 \hline
  & \rot{Reverse Mode}  & \rot{Complexity Proofs} & \rot{Pure} & \rot{Correctness Proofs} & \rot{Tensor Support}  & \rot{Recursion} & \rot{Conditional} \\
  % & Reverse  & Complexity & Pure & Correct & HO Fun. & Tensors & Rec. & Cond. \\
 \hline
 \system{} (This Paper) &
 \supfull & \supfull & \supfull & \supfull  & \supfull  & \supfull & \supfull \\ 
 \hline
 Lantern~\cite{lantern_icfp} & 
 \supfull & \suphalf & \supnone & \supnone & \supfull & \supfull & \supfull \\ 
 \hline
 \dfsmooth{}~\cite{shaikhha2019efficient} 
 & 
 \supnone & \supnone & \supfull & \supnone & \supfull  & \supnone  & \supnone \\ 
 \hline
 \cite{huot2020correctness} &
 \supfull & \supnone & \supfull & \supfull & \supnone & \suphalf & \supfull \\ 
 \hline
 \cite{brunel2019backpropagation} &
 \supfull & \suphalf ? & \supfull & \supfull & \supnone & \supnone & \supnone \\ 
 \hline
 \cite{abadi2019simple} &
 \supfull & \supfull? & \supfull & \supfull & \supnone? & \supfull & \supfull \\ 
 \hline
 \cite{barthe2020versatility} &
 \supnone & \supnone & \supfull & \supfull  & \supnone & \supnone & \supfull \\ 
 \hline
 \cite{pearlmutter2008reverse} &
 \supfull & \supnone  & \supnone & \supnone & \supnone? & \supfull & \supfull \\ 
 \hline
 \cite{Elliott:2018:SEA:3243631.3236765} &
 \supfull & \suphalf & \supfull & \supnone & \supnone? & \supnone & \supnone? \\ 
 \hline
 \cite{sherman2021} & 
 \suphalf or \supnone? & \supnone & \supfull & \supfull? & \supnone? & \supnone & \supnone \\ 
 \hline
 \cite{vytiniotis2019differentiable} &
 \supfull & \suphalf & \supfull & \supnone & \supnone? & \supnone & \supnone? \\ 
 \hline
 \cite{mak2020differential} & 
 \supfull & \suphalf ? & \supfull & \supfull & \supnone & \supnone & \supnone \\ 
 \hline
 \cite{vakar2020reverse} & 
 \supfull & \suphalf ? & \supfull & \supfull & \supnone & \supnone & \supnone \\ 
 \hline
 \cite{Manzyuk2012} & 
 \supnone & \supnone & \supfull & \supfull? & \supnone & \supnone & \supnone \\ 
 \hline 
 \cite{gallagher-sdg} & 
 \supnone & \supnone & \supfull & \supfull ? & \supnone & \supfull & \supfull  \\ 
 \hline
 \end{tabular}
 \caption{Comparison of different differentiable programming frameworks.
 $\supfull$ means the property is verified, and $\supnone$ means its absent from the work.}
 \end{table}

@@

We introduce the general idea of efficient reverse-mode in a functional setting on some examples.
We recall some basics of reverse-mode differentiation in Section~\ref{sec:background}.

\begin{example}[First-order term]
On simple first-order terms, the output code is reminiscent of SSA code~\cite{cytron1989efficient}.
Let us consider the term !let w$_1$ = x$_1$ * x$_2$ in let w$_2$ = w$_1$ * x$_1$ in w$_2$! in the context $\Gamma:=\{x_1,x_2, x_3:\RR\}$.\\
After our (inefficient) reverse-mode transformation, we obtain
\begin{center}
    \begin{tabular}{l}
        !let w$_1$,w$_1'$! = !< x$_1$ * x$_2$, fun (y$_1$,$\ldots$, y$_4$) -> (y$_1$+x$_2$*y$_4$, y$_2$+x$_1$*y$_4$, y$_3$)> in!\\
        !let w$_2$,w$_2'$! = !< w$_1$*x$_1$, fun (y$_1$,$\ldots$, y$_5$) -> w$_1'$!!(y$_1$+w$_1$*y$_5$, y$_2$, y$_3$, y$_4$+x$_1$*y$_5$)> in!\\
        !w$_2'$(0,0,0,0,1)!
    \end{tabular}
\end{center}
The part !(0,0,0,0,1)! comes as an equivalent to initialing the tangent variables in the imperative reverse-mode algorithm.

After some general partial evaluation techniques that will be detailed further in the paper, we obtain

    \begin{center}
            \begin{tabular}{l}
                !let w$_1$ = x$_1$ * x$_2$ in!\\ 
                !let w$_2$ = w$_1$ * x$_1$ in!\\
                !let y$_1$,y$_2$,y$_3$,y$_4$,y$_5$ = 0,0,0,0,1 in!\\
                !let y$_1'$! != y$_1$+w$_1$*y$_5$ in!\\
                !let y$_4'$! != y$_4$+x$_1$*y$_5$ in!\\
                !(y$_1'$+x$_2$*y$_4'$!!, y$_2$+x$_1$*y$_4'$!!, y$_3$)!
            \end{tabular}
    \end{center}   
For the expert viewer, this is equivalent to the SSA form \cite{cytron1989efficient} of what the imperative reverse-mode differentiation of our initial term would be.\\
This term can be further optimized via constant propagation and algebraic simplifications to give
        \begin{center}
            \begin{tabular}{{c}}
                !let w$_1$ = x$_1$ * x$_2$ in!\\ 
                !let w$_2$ = w$_1$ * x$_1$ in!\\
                !(w$_1$+x$_2$*x$_1$, x$_1$*x$_1$, 0)!
            \end{tabular}
        \end{center}
    \end{example}

 \begin{example}[Simple operations on arrays]
    On arrays, three simple operations of interest are the dot-product of two vectors, and the product or sum of the elements of a vector.
    In a functional setting, these can be defined as follows:
\begin{center}
    \begin{tabular}{{r c l}}
        !prod(A)! &:=& !reduce * 1 A! \\
        !sum(A)! &:=& !reduce + 0 A! \\
        !dot(A,B)! &:=& !reduce + 0 (map2 * A B)!     
    \end{tabular}
\end{center}
where !reduce! is a known fold left operator for which the function argument is associative. 
It is notably faster to execute than a fold left as it parallel friendly.

The gradient of each of these construct w.r.t. !A! will be:
\begin{center}
    \begin{tabular}{{r c l}}
        $\nabla_A$!prod(A)! &:=& !map2 * (scanr * 1 (shift1L A)) (shift1R (scanl * 1 A))! \\
        $\nabla_A$!sum(A)! &:=& !map (x -> 1) A!\\
        $\nabla_A$!dot(A,B)! &:=& !B! 
    \end{tabular}
\end{center}
where !scanl! and !scanr! are respectively scan left and scan right. 
These return all the intermediate result of a fold in an array.
!shift1L [v$_1$,$\ldots$,v$_n$] \defeq [v$_2$,$\ldots$,v$_{n}$]! and 
!shift1R [v$_1$,$\ldots$,v$_n$] \defeq [v$_1$,$\ldots$,v$_{n-1}$]! are shift operators.
These gradients are some examples among numerous ones which are usually derived by hand, and are here obtained automatically as special cases of our work.
\end{example}

\subsection{Outline of the paper}

Our main transformation is present on the top of the diagram and presented in Section~\ref{sec:simplediff}. 
It it decomposed 3 steps which are presented in Section~\ref{sec:unf}.
\input{figures/compilation_scheme}

In Section~\ref{sec:background} we recall rudiments of automatic differentiation, forward and reverse-mode differentiation.
Next, in Section~\ref{sec:simplediff} we introduce the source and target languages for differentiation, and our reverse mode transformation. 
In Section~\ref{sec:unf} we introduce a new intermediate representation UNF, and a simple reverse-mode macro on this representation.
We introduce transformations from the source language to UNF and from UNF to the target language.
Next, in Section~\ref{sec:correctness} we give denotational semantics to our languages and show that our decomposition respects the diagram above and that our reverse mode transformation is correct.
In Section~\ref{sec:complexity}, we show that our transformation verifies the cheap gradient principle and extra general optimizations.
In Section~\ref{sec:generalization} we show how to adapt our work for a richer source language. 
Finally, related work, limitations to our approach and future work are presented in Section~\ref{sec:conclusion}.

>>>>

In this paper, we define a compilation scheme for a purely functional, provably correct, and efficient reverse-mode AD. 
To do so, we go start from a simple and general representation, where we can relatively easily prove correctness of reverse-mode.
Then, we define the UNF representation inspired by SSA~\cite{cytron1989efficient} and ANF, that leads to more efficient implementations.

\input{figures/related_work_table}

@@

We introduce the general idea of efficient reverse-mode in a functional setting the next examples.
% We recall some basics of reverse-mode differentiation in Section~\ref{sec:background}.

\begin{example}[First-order term]
% On simple first-order terms, the output code is reminiscent of SSA code~\cite{cytron1989efficient}.
Let us consider the term !let w$_1$ = x$_1$ * x$_2$ in let w$_2$ = w$_1$ * x$_1$ in w$_2$! in the context $\Gamma:=\{x_1,x_2, x_3:\RR\}$.\\
After an (inefficient) reverse-mode transformation, we obtain:
\begin{center}
    \begin{tabular}{l}
        !let w$_1$,w$_1'$! = !< x$_1$ * x$_2$, fun (y$_1$,$\ldots$, y$_4$) -> (y$_1$+x$_2$*y$_4$, y$_2$+x$_1$*y$_4$, y$_3$)> in!\\
        !let w$_2$,w$_2'$! = !< w$_1$*x$_1$, fun (y$_1$,$\ldots$, y$_5$) -> w$_1'$!!(y$_1$+w$_1$*y$_5$, y$_2$, y$_3$, y$_4$+x$_1$*y$_5$)> in!\\
        !w$_2'$(0,0,0,0,1)!
    \end{tabular}
\end{center}
The part !(0,0,0,0,1)! comes as an equivalent to initialing the tangent variables in the imperative reverse-mode algorithm. 
After some general partial evaluation techniques that will be detailed further in the paper, we obtain:

    \begin{center}
            \begin{tabular}{l}
                !let w$_1$ = x$_1$ * x$_2$ in!\\ 
                !let w$_2$ = w$_1$ * x$_1$ in!\\
                !let y$_1$,y$_2$,y$_3$,y$_4$,y$_5$ = 0,0,0,0,1 in!\\
                !let y$_1'$! != y$_1$+w$_1$*y$_5$ in!\\
                !let y$_4'$! != y$_4$+x$_1$*y$_5$ in!\\
                !(y$_1'$+x$_2$*y$_4'$!!, y$_2$+x$_1$*y$_4'$!!, y$_3$)!
            \end{tabular}
    \end{center}   
For the expert viewer, this is equivalent to the SSA form \cite{cytron1989efficient} of what the imperative reverse-mode differentiation of our initial term would be.
This term can be further optimized via constant propagation and algebraic simplifications to give
        \begin{center}
            \begin{tabular}{{c}}
                !let w$_1$ = x$_1$ * x$_2$ in!\\ 
                !let w$_2$ = w$_1$ * x$_1$ in!\\
                !(w$_1$+x$_2$*x$_1$, x$_1$*x$_1$, 0)!
            \end{tabular}
        \end{center}
    \end{example}

 \begin{example}[Simple operations on arrays]
    On arrays, three simple operations of interest are the dot-product of two vectors, and the product or sum of the elements of a vector.
    In a functional setting, these can be defined as follows:
\begin{center}
    \begin{tabular}{{r c l}}
        !prod(A)! &:=& !reduce * 1 A! \\
        !sum(A)! &:=& !reduce + 0 A! \\
        !dot(A,B)! &:=& !reduce + 0 (map2 * A B)!     
    \end{tabular}
\end{center}
where !reduce! is a known fold-left operator for which the function argument is associative. 
It is notably faster to execute than a fold-left as it is parallel friendly.

The gradient of each of these expressions w.r.t. !A! will be:
\begin{center}
    \begin{tabular}{{r c l}}
        $\nabla_A$!prod(A)! &:=& !map2 * (scanr * 1 (shift1L A)) (shift1R (scanl * 1 A))! \\
        $\nabla_A$!sum(A)! &:=& !map (x -> 1) A!\\
        $\nabla_A$!dot(A,B)! &:=& !B! 
    \end{tabular}
\end{center}
where:
\begin{itemize}
\item !scanl! is the scan-left operator that returns all the intermediate results of fold-left,
\item !scanr! is the scan-right operator that returns all the intermediate results of fold-right,
\item !shift1L [v$_1$,$\ldots$,v$_n$]! is the shift-left operator and returns ![v$_2$,$\ldots$,v$_{n}$]!, and 
\item !shift1R [v$_1$,$\ldots$,v$_n$]! is the shift-right operator and returns ![v$_1$,$\ldots$,v$_{n-1}$]!.
\end{itemize}
These gradients are some examples among numerous ones which are usually derived by hand, and are here obtained automatically as special cases of our work.
\end{example}

\input{figures/compilation_scheme}

\subsection{Contributions}
We propose a source-code transformation on a simple purely functional language for purely functional reverse-mode differentiation.
Our transformation is comprised of a compilation scheme that is outlined in Figure~\ref{fig:outline}.
We make the following contributions:

\begin{itemize}[leftmargin=*]
\item We present our work with a simple yet expressive array-based language (with constructs such as !map2! and !reduce!) in Section~\ref{sec:simplediff}. 
We show how to directly compute an efficient reverse-mode AD for the expressions of this program (top of Figure~\ref{fig:outline}).  
Furthermore, we show how to extend our work to a richer language in Section~\ref{sec:generalization}.
\item One of the key insights behind efficient reverse-mode AD is to only consider unary operators. 
Inspired by this insight and following IRs such as SSA and ANF, we introduce a novel IR, UNF (Section~\ref{sec:unf})).
We introduce an alternative and easier to follow compilation pipeline for efficient reverse-mode AD (bottom of Figure~\ref{fig:outline}).
\item We proof the correctness of our transformations (top/bottom parts of Figure~\ref{fig:outline}) by defining the denotational semantics of our languages using multicategories and concategories (Section~\ref{sec:correctness}).
\item We prove the complexity guarantees for the reverse ADed programs. Furthermore, we show a list of optimizations that can further improve the constant factors (Section~\ref{sec:complexity}).
\end{itemize}

Next, we recall rudiments of automatic differentiation, forward and reverse-mode differentiation

% \subsection{Summary of contributions}

% We propose a source-code transformation on a simple purely functional language for purely functional reverse-mode differentiation.
% Our transformation is comprised of a compilation scheme through a novel intermediate representation (IR): unary form (UNF).
% To emphasize the key ideas and for expository purposes, we present our work with a simple yet expressive language in Section~\ref{sec:simplediff}, 
% and show how to extend our work to a richer language in Section~\ref{sec:generalization}.

% We summarize the key ideas of our work. Reverse-mode can be made efficient
% \begin{itemize}
%    \item easily if there are only unary operators
%    \item in a purely functional way
%    \item on expressive constructs such as second-order array operators (e.g. !map2, reduce!) 
%    \item via standard functional optimization techniques
% \end{itemize}

% Our contribution makes these claims precise. In more details, we introduce
% \begin{itemize}
%     \item an expressive source language with array operations, yet restrictive enough for efficient differentiation (\S\ref{sub:sourcelang})
%     \item a method for extending our work and modular approaches to add and prove correct new primitives as needed in practice (\S\ref{sec:generalization})
%     \item a new intermediate representation (IR), UNF, essentially turning every operation into a unary one which inputs and outputs a tuple (\S\ref{sec:unf})
%     \item a transformation between the IR and the main language (\S\ref{sub:transformations to and from UNF})
%     \item a simple reverse-mode transformation on the IR (\S\ref{sub:Simple reverse mode transformation})
%     \item the direct reverse-mode transformation obtained from the source to the target language (\S\ref{sub:Macro for pure reverse mode transformation})
%     \item a denotational semantics of our languages and transformations using multicategories and concategories (\S\ref{sec:correctness})
%     \item correctness of the transformations (\S\ref{sec:correctness})
%     \item complexity guarantees (\S\ref{sec:complexity})
%     \item how efficiency is ensured via standard optimization techniques (\S\ref{sub:Optimizations})
%     \item several additions to the source language (\S\ref{sec:generalization}) TODO:should emphasize that it should be easier for us to add primitives compared to several recent work. give recipe.
% \end{itemize}

% \subsection{Outline of the paper}

% Our main transformation is present on the top of the diagram and presented in Section~\ref{sec:simplediff}. 
% It it decomposed 3 steps which are presented in Section~\ref{sec:unf}.

% \input{figures/compilation_scheme}

% In Section~\ref{sec:background} we recall rudiments of automatic differentiation, forward and reverse-mode differentiation.
% Next, in Section~\ref{sec:simplediff} we introduce the source and target languages for differentiation, and our reverse mode transformation. 
% In Section~\ref{sec:unf} we introduce a new intermediate representation UNF, and a simple reverse-mode macro on this representation.
% We introduce transformations from the source language to UNF and from UNF to the target language.
% Next, in Section~\ref{sec:correctness} we give denotational semantics to our languages and show that our decomposition respects the diagram above and that our reverse mode transformation is correct.
% In Section~\ref{sec:complexity}, we show that our transformation verifies the cheap gradient principle and extra general optimizations.
% In Section~\ref{sec:generalization} we show how to adapt our work for a richer source language. 
% Finally, related work, limitations to our approach and future work are presented in Section~\ref{sec:conclusion}.

### 2021-07-06 rad-icfp-2021/relwork.tex
.

>>>>

\noindent \textbf{Design space}
First, there is a tradeoff to reach between a general expressive language and a domain specific one. The latter usually has more static information and a specific representation that lends itself to better optimizations.
Then, many optimizations performed on AD implementations consist in hand-crafted derivatives for useful operations like matrix-matrix multiplication, dot-product, etc.
They don't seem to arise from theoretical justifications, are error prone, and can hardly fit with more general optimizations.
This makes these systems harder to prove correct. The problem is thus to ensure provable correctness and pureness, while not compromising on efficiency. In addition, we would like something easily provably efficient. 
A real-world implementation based on our work would of course optimize further, potentially using hand-crafted operations as well, but it would be based on solid grounds.
